---
bibliography: references.bib
---

# Exploratory Data Analysis in R - Diagnosing like a Data Doctor {.unnumbered}

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Purpose of workshop

**Exploring a novel data set and produce publication quality tables and reports**

------------------------------------------------------------------------

## Objectives

1.  Load and explore a data set with publication quality tables
2.  Diagnose outliers and missing values in a data set
3.  Prepare an HTML summary report showcasing properties of a data set

------------------------------------------------------------------------

## Required Setup

We first need to prepare our environment with the necessary packages

```{r, results = 'hide', message = FALSE, warning = FALSE}
# Sets the repository to download packages from
options(repos = list(CRAN = "http://cran.rstudio.com/"))

# Sets the number of significant figures to two - e.g., 0.01
options(digits = 2)

# Required package for quick package downloading and loading 
install.packages("pacman")

# Downloads and load required packages
pacman::p_load(dlookr, # Exploratory data analysis
               formattable, # HTML tables from R outputs
               here, # Standardizes paths to data
               kableExtra, # Alternative to formattable
               knitr, # Needed to write HTML reports
               missRanger, # To generate NAs
               tidyverse) # Powerful data wrangling package suite
```

------------------------------------------------------------------------

## Load and Examine a Data Set

-   Load data and view
-   Examine columns and data types
-   Define box plots
-   Describe meta data

We will be using open source data from UArizona researchers for Test, Trace, Treat (T3) efforts offers two clinical diagnostic tests (Antigen, RT-PCR) to determine whether an individual is currently infected with the COVID-19 virus. [@merchant2022]

```{r}
# Let's load a data set from the COVID-19 daily testing data set
dataset <- read.csv(here("Data7_EDA_In_R_Book", "data", "daily_summary.csv")) 

# What does the data look like?
dataset |>
  head() |>
  formattable()
```

------------------------------------------------------------------------

## Diagnose your Data

```{r}
# What are the properties of the data
dataset |>
  diagnose() |>
  formattable()
```

-   `variables`: name of each variable
-   `types`: data type of each variable
-   `missing_count`: number of missing values
-   `missing_percent`: percentage of missing values
-   `unique_count`: number of unique values
-   `unique_rate`: rate of unique value - unique_count / number of observations

------------------------------------------------------------------------

## Summary Statistics of your Data

### Numerical Variables

```{r}
# Summary statistics of our numerical columns
dataset |>
  diagnose_numeric() |>
  formattable()
```

-   `min`: minimum value
-   `Q1`: 1/4 quartile, 25th percentile
-   `mean`: arithmetic mean (average value)
-   `median`: median, 50th percentile
-   `Q3`: 3/4 quartile, 75th percentile
-   `max`: maximum value
-   `zero`: number of observations with the value 0
-   `minus`: number of observations with negative numbers
-   `outlier`: number of outliers

------------------------------------------------------------------------

### Outliers

Values outside of $1.5 * IQR$

![Image Credit: [CÉDRIC SCHERER](https://www.cedricscherer.com/2019/05/17/the-evolution-of-a-ggplot-ep.-1/)](https://d33wubrfki0l68.cloudfront.net/6a759d8217be119e3409d1eb8e6cd78913bcc86f/c1995/img/evol-ggplot/boxplot.png){fig-alt="Boxplot showing the IQR, lower and upper quartiles, median, and outliers"}

------------------------------------------------------------------------

There are several numerical variables that have outliers above, let's see what the data look like with and without them

-   Create a table with columns containing outliers

-   Plot outliers in a box plot and histogram

```{r}
# Table showing outliers
diagnose_outlier(dataset) |>
  filter(outliers_ratio > 0) |>
  formattable()
```

-   `outliers_cnt`: number of outliers
-   `outliers_ratio`: ratio of outliers over all values
-   `outliers_mean`: arithmetic mean (average value) of outlier values
-   `with_mean`: arithmetic mean of all values **including** outliers
-   `without_mean`: arithmetic mean of all values **excluding** outliers

```{r}
# Selecting desired columns 
dataset |>
    plot_outlier()
```

------------------------------------------------------------------------

### Missing Values (NAs)

-   Table showing the extent of NAs in columns containing them
-   Plot showing the frequency of missing values

```{r, output.width = "100%"}
# Create the NA table
dataset |>
  generateNA(p = 0.3) |>
  plot_na_pareto(only_na = TRUE, plot = FALSE) |>
  formattable() # Publishable table
```

```{r}
# Plot the intersect of the columns with the most missing values
# This means that some combinations of columns have missing values in the same row
dataset |>
  generateNA(p = 0.3) |>
  select(test_type, test_result, test_count) |>
  plot_na_intersect(only_na = TRUE) 
```

------------------------------------------------------------------------

### Categorical Variables

```{r}
# Diagnose our categorical columns
dataset |>
  diagnose_category() |> 
  formattable()
```

-   `variables`: category names
-   `levels`: group names within categories
-   `N`: number of observation
-   `freq`: number of observation at group level / number of observation at category level
-   `ratio`: percentage of observation at group level / number of observation at category level
-   `rank`: rank of the occupancy ratio of levels (order in which the groups are in the category)

------------------------------------------------------------------------

## Produce an HTML Summary of a Data Set

```{r}
# Remove the '#' below to reproduce an HTML from an R script. 
#diagnose_web_report(dataset)
```
